Evaluation of a Cache-Oblivious Data Structure [draft-2]

### 2008-06-01 paper/paper.tex
Created final draft version.

\title{Evaluation of a Cache-Oblivious Data Structure [draft-1]}

@@

This model is close enough to the way computers work to be extremely useful in the development and analysis of data structures and algorithms that work well in practice. However, like every model, it is a simplification of reality. One of the simplifications is the assumption that storing and retrieving data in memory is a constant-time operation (i.e. the uniform memory model). Advancements in software and hardware design over the last two de\-cades have caused this assumption to be increasingly detached from reality.

@@

\begin{figure}
\centering
\includegraphics{memhier}
\caption{Schematic depiction of the memory hierarchy}\label{fig-memhier}
\end{figure}

@@

To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed. Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. This has prompted research into new memory models that are more realistic. We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
\begin{list}{}{}
\item \textbf{Cache-unaware} data structures and algorithms are designed with the traditional uniform memory model in mind, without considering the existence of a memory hierarchy, and may therefore not perform optimally in practice.
\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and algorithms are designed to perform optimally in the external memory model (described below) but require parametrization with some or all properties of the cache layout (such as block size or cache memory size).
\item \textbf{Cache-oblivious} data structures and algorithms are designed to perform optimally in the cache-oblivious model (explained below) which does not allow para\-metri\-zation with properties of the cache layout.
\end{list}

Plenty of cache-unaware and cache-aware data structures have been developed and these are also widely used in practice. Although some research has been done on cache-oblivious data structures, at the time of writing there is no evidence of their use in practice. The main goal for this paper is to give some insight in the practical merit of cache-oblivious data structures.

In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious model is of particular interest. We will then describe what our goals for this research project were and how our work relates to previous research.
A large part of the paper will be dedicated to a description of our research methodology. Finally, we will present and discuss our results, and draw a conclusion on the applicability of cache-oblivious data structures.

\section{Previous Work}
The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of these alternatives here.

@@

\subsection{Hierarchical Memory Model}
The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice the memory hierarchy contains more than just two levels. Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one. Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this shortcoming by introducing a hierarchical memory model in which the cost of accessing a value at specific memory addresses is described by a non-decreasing function of these addresses, which means that accessing data at higher addresses can be slower than at lower addresses. This more closely resembles a real memory hierarchy, but assumes that the application has full control over which data is placed where and that data at all locations can be accessed directly, which is usually not the case in practice. As a result, the applicability of their model to the design and evaluation of \emph{practical} algorithms is limited.

@@

The lack of unrealistic requirements combined with the promise of cache-efficiency accross multiple levels of the memory hierarchy, makes it an interesting model for the development of low-maintenance, high-performance data structures for use in real-world applications.

\section{Contributions}
Several cache-oblivious data structures and algorithms have been proposed. Complexity analysis shows that the proposed solutions are asymptotically optimal. However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often preferred that are easier to implement and require less memory).

This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice. To determine if cache-oblivious data structures have practical merit, empirical performance data is required, which is scarce as existing research has focused mainly on theoretical analysis. This paper addresses the question by reporting on the performance of newly implemented (but previously described) cache-oblivious data structures and their traditional counterparts (both cache-aware and cache-unaware data structures).

@@

Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
Later, Bender, Duan, Iacono and Wu simplified this data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for additional operations (finger searches in particular).
This data structure will be explained in detail in Section \ref{sect-bender-set}.
The authors note that a chief advantage of their data structure over the previously described on is that it is less complex, more easily implementable and therefore more suitable for practical use.

@@

We also need to define more accurately what we mean by practical performance. We will execute our test application and measure two properties: primarily, the total running time, and secondarily the memory used. The rationale for selecting these metrics is that if the data structures perform identical functions and plenty of memory is available, the only observable difference in running a program using different data structures will be the run time. Memory use is of secondary interest because in practice memory may be limited, which would preclude the use of certain data structures that require a lot of memory to function.

\subsection{State Space Search}
The test application that we use to gather performance data is an implementation
of a state space search algorithm.
This is a suitable scenario for two reasons.
First, it is commonly used as a practical
component of formal methods for software verification, and therefore good
algorithms are of great practical significance. Second, as we will explain
below, the performance of state space search algorithms depends for a large
part on the performance of the data structures that are used to implement
them; therefore, research into efficient data structures is of particular
interest to this application.

@@

The set of all states reachable by transitions from the initial state is called
the \emph{state space} of a model, and it is the goal of a state space search
algorithm to generate all of these states, in order to check that desired
properties hold in all of them. For this approach it is often required that
the state space be finite (for obvious reasons).

\begin{program}
\begin{verbatim}
Queue queue = new Queue;
Set generated = new Set;
queue.insert(initial_state);
while (!queue.empty()) {
    State state = queue.extract();
    for (State s : successors(state)) {
        if (generated.contains(s) == false) {
            generated.insert(s);
            queue.insert(s);
        }
    }
}
\end{verbatim}
\caption{Pseudo-code for a simple state search algorithm.}
\label{prog-search}
\end{program}

@@

Although variations are possible, our queue operates by a first-in, first-out
principle, meaning that meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps.

From the pseudo-code it is clear that the state space search algorithm does not
accomplish much by itself; instead, the real work is done by the successor
function and the queue and the set data structures. Queues can easily be
implemented efficiently (adding and removing states takes O(1) time).
The efficiency of the successor function depends on the execution model used,
but is typically linear in the number of states produced. In practice,
therefore, the bottleneck of the algorithm is the set data structure.

\subsection{Experimental Framework}
For our experimental framework, we need a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{holzmann2004smc},
a widely used model checking tool.
Spin uses a custom modeling language (called PROMELA) to specify models and is
distributed with a collection of example models that are suitable for our
experiments.

For the execution of these models we will use the NIPS VM \cite{weber2007evm}, a
high-performance virtual machine for state space generation that is easily
embeddable in our framework. Although the NIPS VM only executes bytecode in a
custom format, a PROMELA compiler is also available to generate the required
bytecode from the Spin models \cite{nipsvm}. The NIPS VM is preferred to the
Spin tools because it was designed to be embedded in other programs and as
such is more easily integrated in our framework.

The NIPS VM represents program state as a binary string; the size of the state
depends amongst other on the number of active processes in the program (which
may change over the execution of the program).
A typical state size is in the order of a few hundred bytes.

\subsection{Data Structures}
In the introduction we identified three classes of data structures.
For our evaluation we have implemented a single representative data structure
for each class:
\begin{itemize}
\item \textbf{Cache-unaware}: hash tables. Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
\item \textbf{Cache-aware}: B-trees. The B-tree is the standard choice for storing ordered data on disk, and depends on a page size being selected that corresponds with the size of data blocks that can be efficiently transferred.
\item \textbf{Cache-oblivious}: the data structure proposed by Bender, Duan, Iacono and Wu. Since it provides functionality comparable to that of a B-tree, this seems like a fair candidate for a comparison. For brevity, this data structure will be referred to as a \emph{Bender set}.
\end{itemize}

Cache-oblivious data structures are not yet commonly used in practice and to
our knowledge there are no high-quality implementations publicly available.
The Bender set will therefore have to be implemented from scratch.

In contrast, both hash tables and B-trees are widely used and there are several
high quality implementations available as software libraries. It is, however,
undesirable to use existing libraries for a fair comparison, for two reasons. First, many existing implementations support additional operations (such as removal of elements, locking and synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations. Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not. This may give the existing libraries an unfair advantage.

@@

\subsubsection{Common Implementation Details}
For memory management, the data structures make use of memory mapped files.
The B-tree and hash table implementations use a single file for all data storage;
the Bender tree uses several files (as discussed below) and some main memory.

The use of memory mapped files means that the operating system is in charge of
when and which pages are transferred between main memory and disk storage; the
application assumes no control over this policy. This is in accordance with the
cache-oblivious model, that assumes no control over the cache is possible.
Additionally, memory mapped files bypass the C library's allocation functions
and give the implementer complete control over how data is ordered in (virtual)
memory.

It should be noted that since we only need a limited subset of the functionality
offered by the set data structures for our test application, we did not
implement any operations that were not required to perform our experiments.
However, we did not change the design of the data structures to take advantage
of this. That means that any additional operations could be implemented without
structural changes to our existing implementation.

\subsubsection{Hash Table Implementation}
In its simplest form, a hash table consists of an
index array (the \emph{index}) with a fixed number of slots. A \emph{hash
function} is used to map key values onto slot numbers. If the slot for a value
is empty, it may be inserted there. Queries for the existence of an element in
the hash table similarly see if the queried value is stored at its designated
slot.

When several values are inserted, some values may map to the same slot,
which is problematic if each slot can only store one value. There
are a lot of different ways to resolve this collision problem; we use
\emph{separate chaining}, which means that slots do not store values directly,
but instead each slot stores a pointer to a singly-linked list of values that
map to that slot. This particular implementation technique is well-suited to
the scenario where values may have different sizes (as the slots only need to
store a fixed-size pointer and not a variable-size value) and maintains
relatively good performance when the number of values stored exceeds the size
of the index array \cite{sedgewick1998ac}.

\begin{figure}
\centering
\includegraphics[width=83mm]{hash-table}
\caption{A depiction of a separate-chaining hash table}
\label{fig-hash-table}
\end{figure}

Our hash table implementation uses a fixed size index which must be specified
when creating the hash table, as this simplifies the implementation
considerably. The index is stored at the front of the file, after which values
are simply appended in the order in which they are added. Note that we do not
support removing elements from the hash table, which means we do not have to
deal with holes that would otherwise occur in the stored file.
Figure \ref{fig-hash-table} shows a hash table (with an index size of four,
storing three values) and the way it is stored in consecutive memory.

@@

B-trees are similar to other search tree data structures in the sense that
they store ordered values hierarchically in such a way that each subtree store
values that are close together.
A key property of B-trees is that they do not fix the number of children per
node, but instead organize data in pages of a fixed size, each containing
as many values as will fit. This makes them especially suitable for purposes
of on-disk storage where reading and writing data in a few large blocks is
relatively cheap compared to accessing several smaller blocks.

\begin{figure}
\centering
\includegraphics[width=83mm]{b-tree}
\caption{A depiction of a B-tree}
\label{fig-b-tree}
\end{figure}

B-tree pages are ordered in a tree structure. Figure \ref{fig-b-tree} depicts a
B-tree of height two storing eight values.
In each page the values are stored in lexicographical order, and the values
in the first leaf page are lexicographically smaller than ``John'', the values
in the second leaf page are between ``John'' and ``Philip'' and the values in
the third leaf page are greater than ``Philip''. Note that since the page
size is fixed, not all pages can be completely filled.

@@

\subsubsection{Bender Set Implementation}
\label{sect-bender-set}
The first implementation challenge for the Bender set is that the description
given by Bender et al. assumes that all values stored in the set are of a
fixed size, which is not the case in our experimental framework. To work
around this, we create several separate instances of the Bender set with
different value sizes which are powers of two. When a value is to be inserted,
its size is rounded up to a power of two and it is inserted in the
corresponding set instance. This ensures that the amount of spaces wasted
remains below 50\% while still allowing values of various sizes to be inserted.
In the following description, we will assume that values are of a fixed size.

The implementation of the Bender set consists of two parts: a sparse array
storing the values in order, and a binary tree stored in van Emde Boas
layout (similar to the one proposed by Prokop) that is used to find values
efficiently.
Both the tree and the array may contain special \emph{empty values}.
A depiction of a Bender set storing four values (with a capacity for eight) is
given in Figure \ref{fig-bender-set}, with the window population counts for
the sparse array given on the left (density thresholds not shown) and the index tree for the same array on the right.

@@

A Bender set is created with a capacity $C$ that is a power of two.
Initially, the array stores only empty values and is partitioned into
\emph{windows} on several \emph{levels}.
On the highest level, there is a single window of size $C$. Each subsequent level
has twice as many windows of half that size, and the lowest level has size
$log_2 C$ (rounded up to the next power of 2).
For each level a maximum \emph{density} is chosen, with the lowest level having
density 1, the highest level having density $\delta$, and the density for
the intermediary levels obtained by interpolating linearly between 1 and
$\delta$.

@@

To insert a value, the binary tree is used to find a position in the array
where the value may be inserted (just before its successor). Then, the windows
overlapping this position are considered from bottom to top, and the first
window that can support another element without overflowing is selected, and
then rebalanced (thereby resolving the overflow in all lower level windows).

In the case that all windows overlapping the target position would overflow upon
insertion of the new element, the capacity of the data structure is doubled,
a new top level is created and then the entire array is rebalanced. This is how
the set grows when the number of elements stored comes too close to the storage
capacity. Since changing the capacity causes all values in the array to be moved
and requires that the entire index tree is recreated, this is an expensive
operation; however, it occurs only infrequently.

@@

One thing that is not specified in the paper by Bender et al, is how the
population count for windows is kept. The simplest option is to keep no such
information and simply scan part of the array whenever a population count is
required. Another extreme is to keep population counts for all windows on all
levels, and update these counts whenever values are inserted or moved. In
our experiments a compromise seemed to work best: keep population counts only
for the lowest-level windows, and recompute the counts for higher-level windows
when required. This prevents a lot of recomputation, while keeping the additional
costs of updating low.

@@

In van Emde Boas lay-out, which is defined recursively, a tree with 
$N$ nodes is divided in half vertically, creating a small subtree at the
top and a number of subtrees (approximately $\sqrt{N}$) below. Each subtree
has a size of approximately $\sqrt{N}$ nodes and will be stored in a
contiguous block of memory.

In Figure \ref{fig-van-emde-boas} a binary tree is shown, with the nodes
labeled with their positions according to the van Emde Boas layout.
In this example, only two levels of recursion are needed. If each page
stores three nodes, the path from node 1 to node 12 (or any other leaf node)
requires visits only two pages.

@@

In the index tree used for the Bender set, the leaf nodes store the same values
as the sparse array (including empty values).
Interior nodes store the maximum value of their two children (or an empty value,
if both children store an empty value).
This tree can be used in a similar way as a binary search tree: by examining the
two children of a node, we can determine whether a searched-for value lies in the
left or right subtree.

Note that the \emph{structure} of the tree is static and only changes when the
capacity of the set is increased (in this case, the tree is recreated from scratch).
Of course, the values stored in the tree have to be updated when the corresponding
elements of the array change.

@@

A different way to measure time is using the statistics the kernel collects
of how many seconds the processor spends executing instructions for the process
(both in user space and in kernel space). These values are independent of
what other processes are doing, which makes them more consistent across
multiple executions. However, these values do not account for the time the
system spends waiting (for example, for data to read from disk)
or the time spent by other processes on behalf of the executing process
(for example, the kernel swap daemon on Linux). Since these factors may have
a large effect on the total running time of the algorithm, it is undesirable
to leave these out.

Therefore, we decided to measure wall clock time. To deal with variations
across multiple executions, each experiment is run seven times, and the median
value is used for further analysis.

@@

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux (kernel 2.6) system
with an Intel Xeon E5335 processor (2~GHz, 4~MB cache) and 8~GB of main
memory (no swap file). Although the system is a multiprocessor system
(and the E5335 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

The following models are used for benchmarking:
\begin{itemize}
\item\textbf{Eratosthenes} is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers up to a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only by exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

@@

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the resulting properties of the state space search.
Note that the number of transitions relative to the number of iterations gives an
indication of the ratio of value look-ups (insertions of values that are already
present in the set) and actual insertions, ranging from approximately from 4 to
2.7 look-ups per insertion.

@@

The performance of the Bender set does not seem to depend on the page size,
and as a result the difference between the Bender set and the B-tree is smallest
when the page size is largest. Unfortunately, even then it requires about three
times as much time and a lot more memory.

>>>>

\title{Evaluation of a Cache-Oblivious Data Structure [draft-2]}

@@

This model is close enough to the way computers work to be extremely useful in the development and analysis of data structures and algorithms that work well in practice. However, like every model, it is a simplification of reality. One of the simplifications is the assumption that data can be stored or retrieved at any location in a constant amount of time, which is why we will call this the \emph{uniform memory model}.
Advancements in software and hardware design over the last two de\-cades have caused this assumption to be increasingly detached from reality.

@@

\begin{figure}
\centering
\includegraphics[width=55mm]{memhier}
\caption{Schematic depiction of the memory hierarchy}\label{fig-memhier}
\end{figure}

@@

To conclude: the memory model used in real computer systems is quite a bit more complex than the uniform memory model assumed. Given this reality, the performance of many existing algorithms and data structures can be improved by taking the existence of a memory hierarchy into account. This has prompted research into new memory models that are more realistic. We will consider three classes of data structures and algorithms, depending on the assumptions that are made about the environment in their definition:
\begin{list}{}{}
\item \textbf{Cache-unaware} data structures and algorithms are designed for
the traditional uniform memory model, with no consideration of the existence
of a memory hierarchy.
\item \textbf{Cache-aware} (or \textbf{cache-conscious}) data structures and
algorithms are designed to perform optimally in the external memory model
(described below) but require parametrization with some or all properties of
the cache layout (such as block size or cache memory size).
\item \textbf{Cache-oblivious} data structures and algorithms are designed to
perform optimally in the cache-oblivious model (explained below) which does
not allow para\-metri\-zation with properties of the cache layout.
\end{list}

A great number of cache-unaware and cache-aware data structures has been developed and these are also widely used in practice.
Although some research has been done on cache-oblivious data structures, there is currently no evidence that they are used in practice. Consequently, it is unclear if they are suitable for practical use at all. The main goal for this paper is to give some insight in the practical merit of cache-oblivious data structures.

In the following pages we will give an overview of the different available memory models and explain why the cache-oblivious mod\-el is of particular interest. We will then describe what our goals for this research project were and how our work relates to previous research.
A large part of the paper will be dedicated to a description of our research methodology. Finally, we will present and discuss our results, and draw a conclusion on the applicability of cache-oblivious data structures.

\section{Previous Work}
The cache-oblivious memory model is not the only or the first model that was developed as a more realistic alternative to the uniform memory model. We will briefly describe some of the alternatives.

@@

\subsection{Hierarchical Memory Model}
The external memory model has the limitation that it only describes two levels of storage, while we have seen that in practice the memory hierarchy contains more than just two levels. Even though the external memory model can be used to describe any pair of adjacent levels, a particular algorithm can only be tuned to one.
Aggarwal, Alpern, Chandra and Snir \cite{aggarwal1987mhm} addressed this
shortcoming by introducing a hierarchical memory model in which the cost of
accessing values at different memory addresses is described by a non-decreasing
function of these addresses, which means that accessing data at higher
addresses can be slower than at lower addresses.
This is a very general model that can be applied to the real memory hierarchy,
but it assumes that the application has full control over which data is placed
where, which is usually not the case in practice.
As a result, the applicability of their model to the design and evaluation of
practical algorithms is limited.

@@

The cache-oblivious model is very similar to the real memory hierarchy,
which means that algorithms designed for this model can
easily be implemented in practice. This property, combined with the
promise of cache-efficiency across multiple levels of the memory hierarchy,
makes it a promising model for the development of low-maintenance,
high-performance data structures for use in real-world applications.

\section{Research Goals}
Several cache-oblivious data structures and algorithms have been proposed. Complexity analysis shows that the proposed solutions are asymptotically optimal. However, in software engineering practice we are not only interested in computational complexity, but also in the practical performance of data structures and algorithms. Indeed, many algorithms that have suboptimal computational complexity are actually widely used because they perform well in practice (for example, sorting algorithms like Quicksort and Shell sort) and the converse is true as well: some algorithms, although theoretically sound, are unsuitable for practical use because of greater memory requirements, longer execution time, or difficulty of implementation (for example: linear-time construction of suffix arrays is possible, but in practice slower alternatives are often preferred that are easier to implement and require less memory).

This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice. To determine if cache-oblivious data structures have practical merit, empirical performance data is required, which is scarce, as existing research has focused mainly on theoretical analysis.
This paper addresses the question by reporting on the performance of a newly
implemented (but previously described) cache-oblivious data structure and two
of its traditional counterparts (both cache-aware and cache-unaware data
structures).

@@

Bender, Demaine and Farach-Colton designed a cache-oblivious data structure that supports the same operations as a B-tree \cite{bender2005cob} achieving optimal complexity bounds on search and nearly-optimal bounds on insertion.
Later, Bender, Duan, Iacono and Wu simplified this data structure \cite{bender2004lpc} while preserving the supported operations and complexity bounds and adding support for additional operations (finger searches in particular).
This data structure will be explained in detail in Section \ref{sect-bender-set}.
The authors note that a chief advantage of their data structure over the
previously described one is that it is less complex, more easily implementable
and therefore more suitable for practical use.

@@

We also need to define more accurately what we mean by practical performance.
Our experiments are performed by running a test application (which will be
described in detail below) and measuring two properties: primarily the
execution time, and secondarily the memory in use.
The rationale for selecting these metrics is that if the data structures perform
identical functions and enough memory is available, the only observable
difference in running a program using different data structures will be the
execution time.
Memory use is of secondary interest because in practice memory may be limited,
which would preclude the use of data structures that require a large amount of
memory to function.

\subsection{State Space Search}
The test application that we used to gather performance data implements a state
space search algorithm.
This is a suitable scenario for two reasons.
First, it is commonly used as a practical
component of formal methods for software verification, and therefore good
algorithms are of great practical significance. Second, as we will explain
below, the performance of state space search algorithms depends for a large
part on the performance of the data structures that are used to implement
them; therefore, research into efficient data structures is of particular
interest to this application.

@@

The set of all states reachable by transitions from the initial state is called
the \emph{state space} of a model, and it is the goal of a state space search
algorithm to generate all of these states, in order to check that desired
properties hold in all of them. This approach usually requires the state state
space to be finite, although exhausting the entire state space is not necessary
for our experiments.

\begin{program}[t]
\begin{verbatim}
Queue queue = new Queue;
Set generated = new Set;
queue.insert(initial_state);
while (!queue.empty()) {
    State state = queue.extract();
    for (State s : successors(state)) {
        if (generated.contains(s) == false) {
            generated.insert(s);
            queue.insert(s);
        }
    }
}
\end{verbatim}
\caption{Pseudo-code for a simple state search algorithm.}
\label{prog-search}
\end{program}

@@

Although other behaviour is possible, our queue operates by a first-in,
first-out principle, meaning that all states that are reachable in $N$ steps
from the initial state are expanded before any states that require more than
$N$ steps.

From the pseudo-code it is clear that the state space search algorithm does not
accomplish much by itself; instead, the real work is done by the successor
function and the queue and set data structures. Queues can easily be
implemented efficiently (adding and removing states takes O(1) time).
The efficiency of the successor function depends on the execution model used,
but is typically linear in the number of states produced. In practice,
therefore, the bottleneck of the algorithm is the set data structure.

\subsection{Experimental Framework}
For our experimental framework, we needed a collection of formal models that are
representative of those typically used for formal verification, and a way to
execute them. For the first part, we have looked at Spin \cite{holzmann2004smc},
a widely used model checking tool.
Spin uses a custom modeling language (called PROMELA) to specify models and is
distributed with a collection of example models that are suitable for our
experiments.

For the execution of these models we used the NIPS VM \cite{weber2007evm}, a
high-performance virtual machine for state space generation that is easily
embeddable in our framework. Although the NIPS VM only executes bytecode in a
custom format, a PROMELA compiler is also available to generate the required
bytecode from the Spin models \cite{nipsvm}. The NIPS VM is preferred to the
Spin tools because it was designed to be embedded in other programs and as
such is more easily integrated in our framework.

The NIPS VM represents program state as a binary string; the size of the state
depends (amongst others) on the number of active processes in the program,
which may change over the execution of the program.
A typical state size is in the order of a few hundred bytes.

\subsection{Data Structures}
In the introduction we identified three classes of data structures.
For our evaluation we have implemented a single representative data structure
for each class:
\begin{itemize}
\item \textbf{Cache-unaware}: hash tables. Hash tables are widely used in practice and noted for good performance if the data set fits in main memory, although they have also been used as an index structure for data stored on disk.
\item \textbf{Cache-aware}: B-trees. The B-tree is the standard choice for
storing (ordered) data on disk, and depends on a page size being selected that
corresponds with the size of data blocks that can be efficiently transferred.
\item \textbf{Cache-oblivious}: the data structure proposed by Bender, Duan, Iacono and Wu. Since it provides functionality comparable to that of a B-tree, this seems like a fair candidate for a comparison. For brevity, this data structure will be referred to as a \emph{Bender set}.
\end{itemize}

Cache-oblivious data structures are not yet commonly used in practice and to
our knowledge there are no high-quality implementations publicly available.
The Bender set therefore had to be implemented from scratch.

In contrast, both hash tables and B-trees are widely used and there are several
high quality implementations available as software libraries. It is, however,
undesirable to use existing libraries for a fair comparison, for two reasons. First, many existing implementations support additional operations (such as locking, synchronization, atomic transactions, et cetera) which are not used in our experiments, but which may harm the performance of these implementations. Second, many established libraries have been thoroughly optimized while our newly implemented data structures have not. This may give the existing libraries an unfair advantage.

@@

\subsubsection{Common Implementation Details}
All data structures were implemented in C using a common interface.
For memory management, the data structures make use of a single memory mapped
data region, bypassing the C library's allocation functions and giving the
implementer complete control over how data is arranged in memory.

In principle, this also means the operating system has control over when and
which data pages are transfered from main memory to disk storage and back.
However, our experiments (which were performed on a system without a swap file)
were limited to using main memory only.

It should be noted that since we only need a limited subset of the functionality
offered by the set data structures for our test application, we did not
implement any operations that were not required to perform our experiments.
However, we did not change the design of the data structures to take advantage
of the reduced functionality. That means that additional operations could be
implemented without structural changes to our existing implementation.

\subsubsection{Hash Table Implementation}
In its simplest form, a hash table consists of an
index array (the \emph{index}) with a fixed number of slots. A \emph{hash
function} is used to map values onto slot numbers. If the slot for a value
is empty, it may be inserted there. Queries for the existence of an element in
the hash table similarly see if the queried value is stored at its designated
slot.

When several values are inserted, some values may map to the same slot,
which is problematic if each slot can only store one value. There
are many different ways to resolve this collision problem; we use
\emph{separate chaining}, which means that slots do not store values directly,
but instead each slot stores a pointer to a singly-linked list of values that
map to that slot. This particular implementation technique is well-suited to
the scenario where values may have different sizes (as the slots only need to
store a fixed-size pointer and not a variable-size value) and maintains
relatively good performance when the number of values stored exceeds the size
of the index array \cite{sedgewick1998ac}. Figure \ref{fig-hash-table} shows a
hash table (with an index size of four, storing three values) and the way it
is stored in consecutive memory.

\begin{figure}[t]
\centering
\includegraphics[width=83mm]{hash-table}
\caption{A depiction of a separate-chaining hash table}
\label{fig-hash-table}
\end{figure}

Our hash table implementation uses a fixed size index which must be specified
when creating the hash table, as this simplifies the implementation
considerably. The index is stored at the front of the file, after which values
are simply appended in the order in which they are added. Note that we do not
support removing elements from the hash table, which means we do not have to
deal with holes that would otherwise occur in the stored file.

@@

B-trees are similar to other search tree data structures in the sense that
they store ordered values hierarchically in such a way that every subtree stores
consecutive values.
A key property of B-trees is that (unlike most in-memory tree structures) they
do not fix the number of children per node, but instead organize data in pages
of a fixed size, each containing as many values as will fit.
This makes them especially suitable for purposes of on-disk storage where
reading and writing data in a few large blocks is relatively cheap compared to
accessing several smaller blocks.

\begin{figure}[b]
\centering
\includegraphics[width=83mm]{b-tree}
\caption{A depiction of a B-tree}
\label{fig-b-tree}
\end{figure}

B-tree pages are ordered in a tree structure. Figure \ref{fig-b-tree} depicts a
B-tree of height two storing eight values.
In each page the values are stored in lexicographical order, and the values
in the first leaf page are lexicographically smaller than ``John'', the values
in the second leaf page are between ``John'' and ``Philip'' and the values in
the third leaf page are greater than ``Philip''. Note that since the page
size is fixed, not all pages are completely filled.

@@

\subsubsection{Bender Set Implementation}
\label{sect-bender-set}
The first implementation challenge for the Bender set is that the description
given by Bender et al assumes that all values stored in the set are of a
fixed size, which is not the case in our experimental framework. To work
around this, we create several separate instances of the Bender set with
different value sizes which are powers of two. When a value is to be inserted,
its size is rounded up to a power of two and it is inserted in the
corresponding set instance. This ensures that the amount of space wasted
remains below 50\% while still allowing values of various sizes to be inserted.
The following description will be of a single Bender set instance, and it will
therefore be assumed that values do have a fixed size.

A Bender set has a capacity $C$ that is a power of two. Its implementation
consists of two parts: a sparse array storing the values in order, and a binary
tree stored in van Emde Boas layout that is used to search for values
efficiently.
Both the tree and the array may contain special \emph{empty values}.

Initially, the array stores only empty values and is partitioned into
\emph{windows} on several \emph{levels}.
On the highest level, there is a single window of size $C$. Each subsequent level
has twice as many windows of half that size, and the lowest level has size
$log_2 C$ (rounded up to the next power of 2).
For each level a maximum \emph{density} is chosen, with the lowest level having
density 1, the highest level having density $\delta$, and the density for
the intermediary levels obtained by interpolating linearly between 1 and
$\delta$.

A depiction of a Bender set storing four values (with a capacity for eight) is
given in Figure \ref{fig-bender-set}, with the window population counts for
the sparse array given on the left (density thresholds not shown) and the index
tree on the right.

@@

@@

To insert a value, the index tree is used to find the position of the smallest
existing value that is greater than the new value (i.e. the value's successor);
the new value will be inserted right before its successor.
Then, the windows overlapping the goal position are considered from bottom to
top. The lowest (smallest) window that can support another element without
overflowing is selected, and then rebalanced (thereby resolving the overflow in
the overlapping lower level windows).

If all windows, including the topmost window spanning the entire array,
would overflow upon insertion of the new element, the capacity of the data
structure must be increased to make room for more elements.
When this happens, the capacity is doubled, a new top level is
created and then the entire array is rebalanced.
Since increasing the capacity causes all values in the array to be moved
and the entire index tree to be recreated, this is an expensive operation;
however, it only occurs infrequently.

@@

A detail that is not specified in the paper by Bender et al, is how the
population count for windows is kept. The simplest option is to keep no such
information and simply scan part of the array whenever a population count is
required. Another extreme is to keep population counts for all windows on all
levels, and update these counts whenever values are inserted or moved. In
our experiments a compromise seemed to work best: keep population counts only
for the lowest-level windows, and recompute the counts for higher-level windows
when required. This prevents a lot of recomputation, while keeping the additional
costs of updating low.

@@

In van Emde Boas lay-out, which is defined recursively, a tree with 
$N$ nodes is divided in half vertically, creating a small subtree at the
top and a number of subtrees (approximately $\sqrt{N}$) at the bottom.
Each subtree has a size of approximately $\sqrt{N}$ nodes and is stored in
van Emde Boas layout in a contiguous block of memory; these blocks are then
stored consecutively.

In Figure \ref{fig-van-emde-boas} a binary tree is shown, with the nodes
labeled with their positions according to the van Emde Boas layout.
In this example, only two levels of recursion are needed. If we suppose that
every page stores three nodes, then the highlighted path from node 1 to 12
visits only two pages.

@@

In the index tree used for the Bender set, the leaf nodes store the same values
as the sparse array (including empty values).
Interior nodes store the maximum value of their two children (or an empty value,
if both children store an empty value).
This tree can be used in a similar way as a binary search tree: by examining the
two children of a node, we can determine whether a searched-for value belongs
in the left or right subtree.
See the right side of Figure \ref{fig-bender-set} for an example.

Note that the \emph{structure} of the tree is static and only changes when the
capacity of the set is increased (in which case the tree is recreated from scratch).
Of course, the values stored in the tree have to be updated when the corresponding
elements of the array change.

@@

A different way to measure time is using the statistics the kernel collects
of how many seconds the processor spends executing instructions for the process
(both in user space and in kernel space). These values are independent of
what other processes are doing, which makes them more consistent across
multiple executions. However, these values do not account for the time the
system spends waiting (for example, for data to be read from disk)
or the time spent by other processes on behalf of the executing process
(for example, the kernel swap daemon on Linux). Since these factors may have
a large effect on the total running time of the algorithm, it is undesirable
to leave these out.

Therefore, we decided to measure wall clock time and deal with variations
across multiple executions by running each experiment seven times and using
the median value for further analysis.

@@

\subsubsection{System Configuration}
The experiments where performed on a 64-bit Linux system (kernel version 2.6.18)
with an Intel Xeon E5335 processor (2~GHz, 4~MB cache) and 8~GB of main
memory (no swap file). Although the system is a multiprocessor system
(and the E5335 is a dual-core processor) all code is single-threaded so
only a single core is used when executing tests.

The following models are used for benchmarking:
\begin{itemize}
\item\textbf{Eratosthenes} is a parallel implementation of the sieve
of Eratosthenes which is used to find prime numbers up to a given maximum.
It has a single parameter $N$: only integers from 2 to $N$ (exclusive) are
tested. A new process is created for every prime number found, and as a result
the state space can only be exhaustively searched for relatively small values
of $N$ (e.g. $N < 40$). Because processes are dynamically created, state size
increases during execution.

@@

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the resulting properties of the state space search.
Note that the number of transitions relative to the number of iterations gives an
indication of the ratio of value look-ups (insertions of values that are already
present in the set) and actual insertions, ranging approximately from 4 to
2.7 look-ups per insertion.

@@

The performance of the Bender set does not seem to depend on the page size,
and as a result the difference between the Bender set and the B-tree is smallest
when the page size is largest. Unfortunately, even then it requires about three
times as much time (and 5--6 times as much memory).

The increased memory requirements of the Bender set can in part be explained by
our implementation of variable-length values, which wastes some space by storing
them in fixed-size slots. This overhead should be around 25\% on average.
